Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-11 paper.tex
Trim down ARG section

% Encoding ancestry
% JK: we can say this some other way, this just seemed like the
% most concise way. People may be irritated by being told what
% the definition of a "genome" is, but I think we need to be explicit.
Let us define a genome as the complete set of genetic material
that a child inherits from one parent. Thus, a diploid individual
has two monoploid genomes, one inherited from each parent. Since
each diploid individual lies at the end of two distinct lineages
of descent, they will be represented by \emph{two} places (nodes)
in any genealogical tree. In the tree sequence encoding a
\emph{node} therefore corresponds to a single genome, which
is associated with its creation time (and other optional information),
and recorded in a simple tabular format (Fig.~\ref{fig-ts-example}).
Genetic inheritance between genomes (nodes) % delete?
is defined by edges.
An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
% JK: This is potentially confusing. Changed "genomic interval"
% to "chromosome" interval to avoid talking about "the" genome.
of the contiguous chromosomal segment over which
the child genome inherited genetic material from the parent genome.
Parent and child nodes may correspond to ancestor and descendant
genome separated by many generations.
Critically, edges can span multiple marginal trees along the
genome, and identical node IDs across different trees corresponds
to the same ancestral genome.
For example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral genomes and recording the
contiguous segments over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
See the \nameref{sec-arg} section below for a discussion of this
closely related concept.

% OK, back to tree sequences: what about sequence variation?
The final output of most population genetic simulations is some representation
of sequence variation among the specified samples. For coalescent
simulations, we usually have three steps: (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
combined data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the genetic sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. As well as the huge
difference in storage efficiency, it is often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data stored in the \tskit\ format
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination
stochastic process as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, the interpretation of the term
has shifted away from the original definition as a stochastic
process, to a representation of a \emph{particular} genetic ancestry
using common ancestor and recombination
events~\citep[e.g.][]{minichiello2006mapping}.
More recently, ARGs have essentially come to mean any graph-like representation
of genetic ancestry; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
% Do we include this sentence? It's true, I'm sure: someone new to the field reads
% Iain and Aylwyn's paper, and goes, hmm, this ARG thing sounds interesting,
% let's follow up on that Griffiths reference... Maybe this is part of the
% reason why ARGs have such a fearsome reputation for being difficult:
% there's a basic confusion about what they actually *are*.
Such usage is entirely reasonable, but referring readers back to the
original Griffiths papers does little to clarify.

The tree sequence encoding of ancestry as a set of nodes and edges
as just described, is equivalent to a graph in which edges
are annotated with a set of disjoint genomic intervals.
This is closely related to the Ancestral Recombination Graph (ARG),
a term that is often loosely understood
to mean some representation of genetic
ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
In this sense a succinct tree sequence \emph{is} an ARG.
More technically, an ARG is a particular
Fepresentation of genetic ancestry as a graph based on the
flow of ancestral material through recombination and common
ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
In this sense, tree sequences can be used to \emph{encode}
ARGs without loss of information (although
the default minimal tree sequence encoding is
much more concise; see the \nameref{sec-arg} section).
It should be noted that, despite standardisation
efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
no shared format for interchanging ARG data, and no supporting software
libraries.

@@

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. We do not output the graph directly, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence (see the \nameref{sec-ts} section),
or ``coalescence records'' in the original terminology~\citep{kelleher2016efficient}.
This is far more efficient than outputting the simulated ARG in its entirety.
For a given scaled recombination rate $\rho$
(setting aside the dependency on the sample size $n$)
we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
in an ARG is $O(\rho^2)$,
whereas the size of the tree sequence encoding is
$O(\rho)$~\citep{kelleher2016efficient}.
This difference between a
quadratic and a linear dependency on $\rho$ is profound, and shows why
large simulations cannot output an ARG in practice.

Although by default \msprime\ outputs tree sequences that contain full information
about the genealogical trees, their correlation structure along the genome,
and the ancestral haplotypes on which coalescences occurred, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral haplotypes that were common ancestors but
in which no coalescences occurred, and also information about the precise time
and genomic location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence
(Fig.~\ref{fig-arg}).
One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
and mutational patterns under
the standard coalescent and infinite sites mutation model.
See the Appendix for details on these likelihood calculations.

>>>>

% Encoding ancestry
% JK: we can say this some other way, this just seemed like the
% most concise way. People may be irritated by being told what
% the definition of a "genome" is, but I think we need to be explicit.
Let us define a genome as the complete set of genetic material
that a child inherits from one parent. Thus, a diploid individual
has two (monoploid) genomes, one inherited from each parent. Since
each diploid individual lies at the end of two distinct lineages
of descent, they will be represented by \emph{two} places (nodes)
in any genealogical tree. In the tree sequence encoding a
\emph{node} therefore corresponds to a single genome, which
is associated with its creation time (and other optional information),
and recorded in a simple tabular format (Fig.~\ref{fig-ts-example}).
Genetic inheritance between genomes (nodes) % delete?
is defined by edges.
An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
% JK: This is potentially confusing. Changed "genomic interval"
% to "chromosome" interval to avoid talking about "the" genome.
of the contiguous chromosomal segment over which
the child genome inherited genetic material from the parent genome.
Parent and child nodes may correspond to ancestor and descendant
genomes separated by many generations.
Critically, edges can span multiple marginal trees along the
genome, and identical node IDs across different trees corresponds
to the same ancestral genome.
For example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral genomes and recording the
contiguous segments over which parent-child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
See the \nameref{sec-arg} section below for a discussion of this
closely related concept.

% OK, back to tree sequences: what about sequence variation?
The final output of most population genetic simulations is some representation
of sequence variation among the specified samples. For coalescent
simulations, we usually have three steps: (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
combined data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the genome sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. As well as the huge
difference in storage efficiency, it is often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data stored in the \tskit\ format
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination
stochastic process as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, the interpretation of the term
has shifted away from the original definition as a stochastic
process, to a representation of a particular genetic ancestry
using common ancestor and recombination
events~\citep[e.g.][]{minichiello2006mapping},
to being synonymous with the genetic
ancestry of a sample, without any specific details of how genetic
material disperses~\citep{mathieson2020ancestry}.
Under the latter interpretation,
the tree sequence encoding of genetic ancestry (described above)
clearly \emph{is} an ARG: the nodes and edges
define a graph in which edges are annotated with the
set of disjoint genomic intervals
through which ancestry flows.

% If we interpret an ARG as being a graph describing genetic
% ancestry, then clearly a

% \jkcomment{This paragraph has been copied from the previous
% version of the Tree Sequences section. Update and merge in here
% as appropriate.}
% The tree sequence encoding of ancestry as a set of nodes and edges
% as just described, is equivalent to a graph in which edges
% are annotated with a set of disjoint genomic intervals.
% This is closely related to the Ancestral Recombination Graph (ARG),
% a term that is often loosely understood
% to mean some representation of genetic
% ancestry as a graph~\citep[e.g.][]{mathieson2020ancestry}.
% In this sense a succinct tree sequence \emph{is} an ARG.
% More technically, an ARG is a particular
% Fepresentation of genetic ancestry as a graph based on the
% flow of ancestral material through recombination and common
% ancestor events~\citep{griffiths1991two,griffiths1997ancestral}.
% In this sense, tree sequences can be used to \emph{encode}
% ARGs without loss of information (although
% the default minimal tree sequence encoding is
% much more concise; see the \nameref{sec-arg} section).
% It should be noted that, despite standardisation
% efforts~\citep{cardona2008extended,mcgill2013graphml}, there is
% no shared format for interchanging ARG data, and no supporting software
% libraries.

@@

For our purposes, an ARG is a realisation of the coalescent with
recombination, in the Griffiths (little ARG) sense.
As described in detail by~\cite{kelleher2016efficient}, Hudson's algorithm
works by dynamically generating and traversing a little ARG.
The graph is not explicitly represented in memory, but is partially
present through the extant lineages and the ancestral material they carry
over time. We do not output the graph directly, but
rather store the information required to recover the genealogical
history as nodes and edges in a tree sequence.
This is far more efficient than outputting the simulated ARG in its entirety.
For a given scaled recombination rate $\rho$
(setting aside the dependency on the sample size $n$)
we know from Eq.~\eqref{eqn-hsw-bound} that the number of nodes
in an ARG is $O(\rho^2)$,
whereas the size of the tree sequence encoding is
$O(\rho)$~\citep{kelleher2016efficient}.
This difference between a
quadratic and a linear dependency on $\rho$ is profound, and shows why
large simulations cannot output an ARG in practice.

Although by default \msprime\ outputs tree sequences that contain full information
about the genealogical trees, their correlation structure along the chromosome,
and the ancestral genomes on which coalescences occurred, some information
is lost in this mapping down from ARG space to the minimal tree sequence
form. In particular, we lose
information about ancestral genomes that were common ancestors but
in which no coalescences occurred, and also information about the precise time
and chromosomal location of recombination events. In most cases, such
information is of little relevance as it is in principle unknowable,
but there are occasions such as visualisation or computing likelihoods (see
below) in which is useful. We therefore provide
the \texttt{record\_full\_arg} option in \msprime\
to store a representation of the complete ARG traversed during simulation.
This is done by storing extra nodes (marked with specific flags, so they
can be easily identified later) and edges in the tree sequence
(Fig.~\ref{fig-arg}).
One situation in which a record of the full ARG is necessary is when we
wish to compute likelihoods during inference.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as
Metropolis-Hastings~\citep{kuhner2000maximum,nielsen2000estimation,wang2008bayesian}.
We provide functions to compute the likelihood of ARG realisations
and mutational patterns under
the standard coalescent and infinite sites mutation model.
See the Appendix for details on these likelihood calculations.
